---
title: "Benchmarking and Reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking and Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`expranno` aims to maximize annotation coverage, but public-facing
analysis also needs provenance and benchmarkable behavior.

This article focuses on three things:

1. comparing annotation engines on the same input
2. validating exact matches against a truth table
3. recording backend provenance for reproducibility
4. keeping a session-level record of the software environment

```{r}
library(expranno)
demo <- example_expranno_data("human")
truth <- example_annotation_truth("human")
```

## Benchmark annotation engines

Use `benchmark_annotation_engines()` when you want to compare
`biomart`, `orgdb`, `ensdb`, and `hybrid` on exactly the same data.

```{r}
benchmark <- benchmark_annotation_engines(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  engines = c("none"),
  verbose = FALSE
)

benchmark$summary
```

In a real run with optional backends installed, a more informative call
looks like this:

```{r, eval = FALSE}
benchmark <- benchmark_annotation_engines(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  engines = c("biomart", "orgdb", "ensdb", "hybrid"),
  biomart_version = 102
)
```

The benchmark output helps answer practical questions:

- does `hybrid` improve `symbol` coverage over `biomart` alone?
- which engine contributes most to `gene_name` coverage?
- how many ambiguous gene-fields appear under each engine?

## Validate against truth

Coverage is useful, but a public package also needs a way to ask
whether recovered labels are actually correct for a known subset.

```{r}
validation <- validate_annotation_engines(
  expr = demo$expr,
  meta = demo$meta,
  truth = truth,
  species = "human",
  engines = "none",
  fields = "symbol",
  verbose = FALSE
)

validation$summary
```

In a production run, the same function can be used to compare `biomart`,
`orgdb`, `ensdb`, and `hybrid` against a curated truth set from the same
species.

The package-level helpers now make that easy to reproduce in examples:

- `example_annotation_truth("human")`
- `example_annotation_truth("mouse")`

## Provenance from annotation

`annotate_expr()` now records backend provenance and ambiguity directly.

```{r}
annotation <- annotate_expr(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none",
  verbose = FALSE
)

annotation$provenance
annotation$ambiguity_report
```

For production runs with `annotation_engine = "hybrid"`, the provenance
table becomes much more informative. It records which backend ran, what
species and dataset were used, which Ensembl release was targeted, and
the annotation date.

## Stable files from the full wrapper

`run_expranno()` writes reproducibility-oriented side files by default.

```{r}
outdir <- tempfile("expranno-benchmark-")
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

result <- run_expranno(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none",
  output_dir = outdir,
  run_deconvolution = FALSE,
  run_signature = FALSE,
  run_benchmark = TRUE,
  benchmark_engines = c("none"),
  verbose = FALSE
)

basename(result$files)
```

The wrapper now keeps:

- `annotation_report.csv`
- `annotation_ambiguity.csv`
- `annotation_provenance.csv`
- `annotation_benchmark_summary.csv`
- `annotation_benchmark_coverage.csv`
- `annotation_validation_summary.csv`
- `annotation_validation_detail.csv`
- `session_info.txt`

These files make it easier to re-run the same workflow later and explain
which annotation sources were active in a given result directory.

For a more reproducible machine-level setup, `expranno` also ships:

- `inst/scripts/install_optional_backends.R`
- `inst/docker/Dockerfile`
