---
title: "Case Study: Mouse RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Mouse RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(expranno)
demo <- example_expranno_data("mouse")
mock_expr_anno <- data.frame(
  gene_id_raw = c("ENSMUSG00000059552.8", "ENSMUSG00000064341.1", "ENSMUSG00000017167.15"),
  gene_id = c("ENSMUSG00000059552", "ENSMUSG00000064341", "ENSMUSG00000017167"),
  symbol = c("Trp53", "mt-Nd1", "Egfr"),
  gene_name = c(
    "transformation related protein 53",
    "mitochondrially encoded NADH dehydrogenase 1",
    "epidermal growth factor receptor"
  ),
  annotation_status = c("annotated", "annotated", "annotated"),
  stringsAsFactors = FALSE
)
mock_report <- data.frame(
  field = c("symbol", "gene_name", "entrez_id", "biotype"),
  annotation_rate = c(1.00, 1.00, 0.67, 1.00)
)
mock_provenance <- data.frame(
  source = c("biomart", "orgdb", "ensdb"),
  backend_release = c("Ensembl v102", "org.Mm.eg.db 3.22.0", "EnsDb.Mmusculus.v79 2.99.0"),
  annotation_date = c("2026-04-03", "2026-04-03", "2026-04-03"),
  status = c("ok", "ok", "ok"),
  stringsAsFactors = FALSE
)
```

## Goal

This case study shows the same workflow for mouse RNA-seq data. The key
difference is that the species and mouse annotation backends need to be
set correctly.

## Step 1: Input tables

```{r}
demo$expr
demo$meta
```

## Step 2: Validate the input contract

```{r}
validate_expr(demo$expr)
validate_meta(demo$meta)
```

## Step 3: Choose an annotation engine

For mouse data, the recommended production choice is:

- `annotation_engine = "hybrid"`

That setting tries:

1. `biomaRt`
2. `org.Mm.eg.db`
3. optional `EnsDb.Mmusculus.v79`

This matters because mouse Ensembl IDs begin with `ENSMUSG`, and the
symbol mapping must come from mouse-specific resources. As in the human
article, a small mock output is shown below so the docs can display a
nonzero coverage example without heavy annotation backends.

```{r}
anno <- annotate_expr(
  expr = demo$expr,
  meta = demo$meta,
  species = "mouse",
  annotation_engine = "none"
)

anno$expr_anno
```

## Step 4: Inspect `expr_anno.csv`

```{r}
mock_expr_anno
```

For mouse data, this is where you check that the normalized IDs still
follow the expected `ENSMUSG...` pattern and that downstream annotation
fields were filled when a real backend is used.

## Step 5: Inspect the annotation report

```{r}
mock_report
```

## Step 6: Inspect provenance

```{r}
mock_provenance
```

## Step 7: How to read annotation coverage

For mouse annotation, the report helps answer three questions quickly:

1. was `symbol` coverage high enough for downstream tools?
2. were `gene_name` and `entrez_id` filled as expected?
3. do the fields look consistent with mouse-specific resources?

In this mock example, the key symbol and gene name fields are fully
covered, while `entrez_id` is only partially covered. That is a useful
signal: the annotation is good enough for symbol-based Deconvolution and
Signature workflows, but not every identifier family is equally complete.

If the report is unexpectedly weak, check:

- `species = "mouse"`
- mouse annotation backends are installed
- the input IDs are real mouse Ensembl gene IDs

## Step 8: Merge annotated expression with metadata

```{r}
merged <- merge_expr_meta(
  expr_anno = anno$expr_anno,
  meta = anno$meta_checked
)

merged
```

## Step 9: Inspect a simple sample summary

```{r}
aggregate(expression ~ sample + group, data = merged, FUN = mean)
```

## Step 10: Prepare signature inputs

```{r}
geneset_path <- system.file("extdata", "hallmark_demo.gmt", package = "expranno")
read_genesets(geneset_path)
```

## Mouse workflow takeaway

The structure is the same as the human case study, but the species must
be set to `"mouse"` and the annotation backend needs mouse-compatible
resources. The annotation report and provenance table are the fastest way
to confirm that the species-specific mapping worked and which release was
used.
