---
title: "Case Study: Human RNA-seq Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Human RNA-seq Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(expranno)
demo <- example_expranno_data("human")
mock_expr_anno <- data.frame(
  gene_id_raw = c("ENSG00000141510.17", "ENSG00000146648.18", "ENSG00000157764.15"),
  gene_id = c("ENSG00000141510", "ENSG00000146648", "ENSG00000157764"),
  symbol = c("TP53", "EGFR", "BRAF"),
  gene_name = c(
    "tumor protein p53",
    "epidermal growth factor receptor",
    "B-Raf proto-oncogene, serine/threonine kinase"
  ),
  annotation_status = c("annotated", "annotated", "annotated"),
  stringsAsFactors = FALSE
)
mock_report <- data.frame(
  field = c("symbol", "gene_name", "entrez_id", "biotype"),
  annotation_rate = c(1.00, 1.00, 0.67, 1.00)
)
mock_provenance <- data.frame(
  source = c("biomart", "orgdb", "ensdb"),
  backend_release = c("Ensembl v102", "org.Hs.eg.db 3.22.0", "EnsDb.Hsapiens.v86 2.99.0"),
  annotation_date = c("2026-04-03", "2026-04-03", "2026-04-03"),
  status = c("ok", "ok", "ok"),
  stringsAsFactors = FALSE
)
```

## Goal

This case study shows a simple human RNA-seq workflow in `expranno`,
starting from `expr` and `meta`, then moving through annotation,
integration, and output generation.

## Step 1: Input tables

```{r}
demo$expr
demo$meta
```

## Step 2: Validate the input contract

```{r}
validate_expr(demo$expr)
validate_meta(demo$meta)
```

## Step 3: Choose an annotation engine

For human data, the recommended production choice is:

- `annotation_preset = "human_tpm_v102"` for TPM-like input
- `annotation_preset = "human_count_v102"` for count-like input

That setting tries:

1. `biomaRt`
2. `org.Hs.eg.db`
3. optional `EnsDb.Hsapiens.v86`

For the website build, we use `"none"` so the article stays lightweight.
To still show what a successful annotation pass looks like, the article
also includes a small mock output table and coverage report.

```{r}
anno <- annotate_expr(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none"
)

anno$expr_anno
```

## Step 4: Inspect `expr_anno.csv`

In a real run, this table is what gets written to `expr_anno.csv`.
Below is a compact mock excerpt that shows the kind of annotated output
you should expect from a successful human run with `"hybrid"`.

```{r}
mock_expr_anno
```

The main columns to check are:

- `gene_id_raw`
- `gene_id`
- `symbol`
- `gene_name`
- `annotation_status`

## Step 5: Inspect the annotation report

`expranno` returns a compact report table that summarizes how many genes
received each annotation field.

```{r}
mock_report
```

## Step 6: Inspect provenance

In a real hybrid run, `annotation_provenance.csv` records which backend
ran and which release was used.

```{r}
mock_provenance
```

## Step 7: How to read annotation coverage

The `annotation_rate` column is the proportion of genes with a non-missing
value for each field in the annotated gene table.

The main practical checks are:

1. `symbol` coverage should be high enough for downstream symbol-based tools
2. `gene_name` coverage should usually track with symbol coverage
3. `entrez_id` can be a little lower, but large gaps still deserve attention
4. low rates suggest rerunning with `"hybrid"` and checking installed backends

In this mock example, `symbol` and `gene_name` are fully covered, while
`entrez_id` coverage is partial. That pattern is often acceptable if the
next steps mainly depend on symbols, but it would still be worth noting in
the analysis record.

## Step 8: Merge annotated expression with metadata

```{r}
merged <- merge_expr_meta(
  expr_anno = anno$expr_anno,
  meta = anno$meta_checked
)

merged
```

## Step 9: Inspect a simple sample summary

```{r}
aggregate(expression ~ sample + group, data = merged, FUN = mean)
```

## Step 10: Run the standard wrapper

```{r}
out <- run_expranno(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none",
  output_dir = tempdir(),
  run_deconvolution = FALSE,
  run_signature = FALSE
)

out
```

## Human workflow takeaway

For a real human analysis, the usual sequence is:

1. annotate with a fixed preset such as `"human_tpm_v102"`
2. inspect `expr_anno.csv`
3. inspect `annotation_report.csv`
4. inspect `annotation_provenance.csv`
5. merge with `meta`
6. run deconvolution
7. run GSVA or ssGSEA
