---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`expranno` is designed for a narrow input contract.

- `expr`: first column is `gene_id`
- `meta`: first column is `sample`

```{r}
library(expranno)

demo <- example_expranno_data("human")
demo$expr
demo$meta
```

The lightest end-to-end run uses `annotation_engine = "none"`. This keeps the
example independent of optional annotation backends.

```{r}
result <- run_expranno(
  expr = demo$expr,
  meta = demo$meta,
  annotation_engine = "none",
  run_deconvolution = FALSE,
  run_signature = FALSE
)
```

The annotation layer returns an annotated table and a compact report.

```{r}
result$annotation$expr_anno
result$annotation$report
```

The merged output is a long-format table that joins expression values to sample
metadata.

```{r}
head(result$expr_meta_merged)
```

For real annotation, switch `annotation_engine` to `"hybrid"`, `"biomart"`, or
`"orgdb"` after installing the required optional packages.

