---
title: "Getting Started with expranno"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with expranno}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(expranno)
demo <- example_expranno_data()
```

`expranno` expects:

- an expression matrix-like data frame with `gene_id` in the first column
- sample columns after `gene_id`
- a metadata table with `sample` in the first column

<style>
.mermaid {
  margin: 1.5rem auto 2rem auto;
  max-width: 980px;
  padding: 0.75rem;
  background: #f7f9fc;
  border: 1px solid #d9e4ef;
  border-radius: 16px;
}
</style>
<pre class="mermaid">
flowchart TB
  subgraph inputs["Inputs"]
    expr["expr<br/>gene x sample table"]
    meta["meta<br/>sample x metadata table"]
  end

  subgraph core["Core workflow"]
    validate["Validate<br/>check required columns"]
    annotate["Annotation<br/>normalize Ensembl IDs<br/>hybrid human or mouse mapping"]
    report["Coverage report<br/>annotation_rate by field"]
    merge["Merge<br/>combine expr_anno and meta"]
  end

  subgraph outputs["Saved outputs"]
    expranno["expr_anno.csv"]
    merged["expr_meta_merged.csv"]
  end

  subgraph downstream["Downstream analyses"]
    deconv["Deconvolution<br/>immunedeconv<br/>one CSV per method"]
    signature["Signature<br/>GSVA or ssGSEA<br/>one CSV per method"]
  end

  expr --> validate
  meta --> validate
  validate --> annotate
  annotate --> expranno
  annotate --> report
  expranno --> merge
  meta --> merge
  merge --> merged
  merged --> deconv
  merged --> signature

  classDef input fill:#eef5fb,stroke:#315c86,color:#17324d,stroke-width:1.5px;
  classDef process fill:#ffffff,stroke:#315c86,color:#17324d,stroke-width:1.5px;
  classDef output fill:#eaf6ef,stroke:#3c7a57,color:#17324d,stroke-width:1.5px;

  class expr,meta input;
  class validate,annotate,report,merge,deconv,signature process;
  class expranno,merged output;
</pre>
<script src="../deps/mermaid/mermaid.min.js"></script>
<script>
if (window.mermaid) {
  window.mermaid.initialize({
    startOnLoad: true,
    securityLevel: "loose",
    theme: "base",
    themeVariables: {
      primaryColor: "#ffffff",
      primaryTextColor: "#17324d",
      primaryBorderColor: "#315c86",
      lineColor: "#315c86",
      secondaryColor: "#eef5fb",
      tertiaryColor: "#eaf6ef",
      clusterBkg: "#f7f9fc",
      clusterBorder: "#cfddeb",
      fontFamily: "Lato, Arial, sans-serif"
    },
    flowchart: {
      curve: "basis",
      htmlLabels: true
    }
  });
  window.mermaid.run({ querySelector: ".mermaid" });
}
</script>

## The workflow at a glance

The workflow is easiest to read in four checkpoints:

1. validate the input contract
2. annotate genes and inspect coverage
3. merge expression with metadata
4. run Deconvolution or Signature analyses from the merged table

## Pick a species

Use `species = "human"` for `ENSG...` input IDs and `species = "mouse"`
for `ENSMUSG...` input IDs. If IDs are clean Ensembl IDs, `species = "auto"`
can infer this for you.

```{r}
human_demo <- example_expranno_data("human")
mouse_demo <- example_expranno_data("mouse")

head(human_demo$expr$gene_id)
head(mouse_demo$expr$gene_id)
```

## Pick an annotation engine

`expranno` exposes five annotation modes:

- `"hybrid"`: recommended production mode
- `"biomart"`: use only `biomaRt`
- `"orgdb"`: use only `org.Hs.eg.db` or `org.Mm.eg.db`
- `"ensdb"`: use only `EnsDb`
- `"none"`: skip annotation lookup and keep the normalized IDs only

`"hybrid"` is the default because it is the most coverage-oriented mode.
It uses a cascade:

1. `biomaRt`
2. species-specific `orgdb`
3. species-specific `EnsDb`

For lightweight docs and tests, the examples below use `"none"`.

```{r}
annotated <- annotate_expr(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none"
)

merged <- merge_expr_meta(
  expr_anno = annotated$expr_anno,
  meta = annotated$meta_checked
)
```

The all-in-one wrapper is `run_expranno()`.

```{r}
result <- run_expranno(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "none",
  output_dir = tempdir(),
  run_deconvolution = FALSE,
  run_signature = FALSE
)

result
```

## Control duplicate symbols and score kernels

Downstream methods usually operate on gene symbols, so duplicated symbols
have to be collapsed. `expranno` now makes that rule explicit.

- `expr_scale = "count"` usually pairs with `duplicate_strategy = "sum"`
- `expr_scale = "abundance"` or `"log"` usually pairs with `duplicate_strategy = "mean"`

For signature scoring, `signature_kcdf` should also match the expression
scale:

- use `"Poisson"` for count-like input
- use `"Gaussian"` for continuous log-scale input
- use `"auto"` if you want GSVA to choose

## Method-specific deconvolution arguments

Some `immunedeconv` methods require extra arguments. In particular,
`timer` and `consensus_tme` need an `indications` vector with one tumor
type per sample.

```{r, eval = FALSE}
run_expranno(
  expr = demo$expr,
  meta = demo$meta,
  species = "human",
  annotation_engine = "hybrid",
  output_dir = "results",
  run_deconvolution = TRUE,
  deconv_args = list(indications = c("SKCM", "SKCM")),
  run_signature = TRUE,
  geneset_file = "hallmark.gmt",
  signature_method = "both",
  signature_kcdf = "Gaussian"
)
```

## What to inspect after a run

For a production analysis with `"hybrid"`, the first things to check are:

- `expr_anno.csv`
- the annotation coverage report from `annotate_expr()`
- `expr_meta_merged.csv`

If `symbol` coverage is weak, downstream Deconvolution and Signature
methods will usually be the first places where the problem appears.
