---
title: "Guide: Annotation Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide: Annotation Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`annotate_expr()` supports four modes.

- `"hybrid"`: `biomaRt` first, `OrgDb` fallback
- `"biomart"`: `biomaRt` only
- `"orgdb"`: `OrgDb` only
- `"none"`: no external annotation

Use `"none"` for tests, documentation, and environments where optional
annotation packages are not installed. Use `"hybrid"` for normal production
work when both backends are available.

```{r}
library(expranno)

demo <- example_expranno_data("mouse")
anno <- annotate_expr(demo$expr, annotation_engine = "none")
anno$expr_anno
anno$report
```

The annotation columns are:

- `gene_id_raw`
- `gene_id`
- `symbol`
- `gene_name`
- `annotation_source`
- `annotation_status`

When `output_dir` is supplied, `annotate_expr()` writes:

- `expr_anno.csv`
- `annotation_report.csv`

