---
title: "Guide: Analysis Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide: Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`expranno` treats deconvolution and signature scoring as optional downstream
steps.

## Deconvolution

`run_cell_deconvolution()` expects an annotated matrix and uses the `symbol`
column to build a gene-by-sample matrix. By default it uses a small method set
to keep behavior predictable:

- human: `quantiseq`, `mcp_counter`, `xcell`
- mouse: `mmcp_counter`

The function only runs when `immunedeconv` is installed.

## Signature scoring

`run_signature_analysis()` accepts either:

- a named list of gene sets
- a GMT file path

The package does not bundle large gene-set resources. That is intentional.
Large signature databases should remain external so the package stays small.

```{r}
library(expranno)

demo <- example_expranno_data("human")
anno <- annotate_expr(demo$expr, annotation_engine = "none")

genesets <- list(
  toy_signal = c("TP53", "EGFR"),
  toy_dna_repair = c("BRCA1", "TP53")
)

genesets
```

This package example stops before scoring because `annotation_engine = "none"`
does not produce usable symbols. In a real run, use a real annotation backend
first, then call `run_signature_analysis()` or `run_cell_deconvolution()`.

