---
title: "Preset Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preset Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`expranno` ships with fixed annotation presets so repeated human and
mouse workflows can share the same release, species, and version
stripping rules.

```{r}
library(expranno)

list_annotation_presets()
```

## How to read the preset table

- `recommended_input` tells you which expression scale the preset is
  designed around.
- `biomart_version` fixes the targeted Ensembl release.
- `fallback_order` shows the human or mouse backend cascade used by the
  hybrid engine.
- `bundled_truth` tells you which built-in truth table pairs naturally
  with the preset for validation examples.

## Typical choices

For human TPM workflows:

```{r, eval = FALSE}
result <- run_expranno(
  expr = expr,
  meta = meta,
  annotation_preset = "human_tpm_v102",
  expr_scale = "abundance",
  duplicate_strategy = "mean",
  run_deconvolution = TRUE,
  run_signature = TRUE,
  geneset_file = "hallmark.gmt",
  signature_kcdf = "Gaussian"
)
```

For mouse count workflows:

```{r, eval = FALSE}
result <- run_expranno(
  expr = expr,
  meta = meta,
  annotation_preset = "mouse_count_v102",
  expr_scale = "count",
  duplicate_strategy = "sum",
  run_deconvolution = TRUE,
  run_signature = TRUE,
  geneset_file = "hallmark.gmt",
  signature_kcdf = "Poisson"
)
```

## Built-in truth resources

The package now bundles small human and mouse truth tables that match the
example Ensembl IDs and are useful for demos, tests, and reproducible
validation tutorials.

```{r}
example_annotation_truth("human")
example_annotation_truth("mouse")
```

These are intentionally small. They are not meant to replace a curated
project-specific truth set, but they make the validation workflow easy to
reproduce across machines and in documentation.
